ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data

### 2021-01-03 manuscript/supplement.tex
manuscript: update to latest text, scripts, figures and results data

\documentclass[authoryear,preprint,11pt]{scrartcl}
\addtokomafont{disposition}{\rmfamily}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{url}
\usepackage{epsfig}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

@@

\begin{figure}[tbp]
 \includegraphics[width=.7\linewidth]{figs/Event_space.pdf}\\
 \caption{
  It depicts single cell events at a particular genomic site, defined by combinations of the possible true alternative nucleotide frequencies in single cell sequencing ($\theta_s$) and bulk sequencing samples ($\theta_b$) in our probabilistic model.
  The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
  A tabular representation of the same event definitions is given in Table~\ref{tab:Events} below (this is a larger copy of Figure~\ref{fig:prosolo_alt-calling}D in the main manuscript, reproduced  here for easier cross-reference with that table).\footnotesize\newline
  ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference
  }
 \label{fig:event-space}
\end{figure}

\begin{center}
 \begin{table}[tbp]
  \caption{
   Single cell events at a particular genomic site are defined by combinations of the possible true alternative nucleotide frequencies in single cell sequencing ($\theta_s$) and bulk sequencing samples ($\theta_b$) in our probabilistic model.
   A graphical representation of the same event definitions is given in Figure~\ref{fig:event-space}.
  }
  \label{tab:Events}
  \renewcommand{\arraystretch}{0.9}

  \begin{tabular}{p{22ex}p{12ex}p{12ex}p{33ex}}
   \toprule
   single cell event & $\theta_s(x)$   & $\theta_b(x)$ & implicit single cell genotype \\
   \toprule
   $\boldsymbol{E}_{\text{hom ref}}$      & $0$             & $[0,\frac12)$ & homozygous ref                \\
       \midrule
   $\boldsymbol{E}_{\text{ADO to ref}}$   & $0$             & $[\frac12,1)$ & heterozygous                  \\
         \midrule
   $\boldsymbol{E}_{\text{err ref}}$      & $\{0,\frac12\}$ & $1$           & homozygous alt                \\
         \midrule
   $\boldsymbol{E}_{\text{err alt}}$      & $\{\frac12,1\}$ & $0$           & homozygous ref                \\
         \midrule
   $\boldsymbol{E}_{\text{het}}$          & $\frac12$       & $(0,1)$       & heterozygous                  \\
         \midrule
   $\boldsymbol{E}_{\text{ADO to alt}}$   & $1$             & $(0,\frac12]$ & heterozygous                  \\
       \midrule
   $\boldsymbol{E}_{\text{hom alt}}$      & $1$             & $(\frac12,1]$ & homozygous alt                \\
   \bottomrule
   \multicolumn{4}{l}{\footnotesize ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous,}\\
   \multicolumn{4}{l}{\footnotesize hom -- homozygous, ref -- reference} \\
   \bottomrule
  \end{tabular}
 \end{table}
\end{center}

Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell and the bulk sequencing sample, given the respective data.
All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequencies $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
We then define mutually exclusive subspaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}$, providing a set of events $\mathfrak{E}$.
Assuming that there is no informative prior information about different allele frequencies (i.e. assuming a flat prior across all possible allele frequencies in both samples), we can then calculate the likelihood of such an event as the product of the likelihoods of the allele frequency ranges in the two samples.
For $\boldsymbol{E}_{\text{ADO to ref}}$---and using Equation \ref{eq:thetas-likelihood} to calculate the likelihoods of the allele frequency ranges---this for example becomes:

@@

\begin{figure}[!tpb]
 \begin{minipage}{.99\linewidth}
  \includegraphics[width=.99\linewidth]{figs/Dong2017/Dong2017_coverage_dist.pdf}
 \end{minipage} \\
 \begin{minipage}{.99\linewidth}
  \includegraphics[width=.99\linewidth]{figs/Laehnemann2017/Laehnemann2017_coverage_dist.pdf}
 \end{minipage}
 \caption{
 Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples.
 Panel \textbf{A} shows the coverage of the single cells (IL-11, IL-12), the kindred ground truth clone (IL-1c) and the more distant background bulk (Clones C1, C2, and C3 joined into one sample) of the dataset from \cite{dong_accurate_2017}.
 Panel \textbf{B} shows the coverage of the single cells (PAG1, PAG2, PAG5, PAG9, PAG10) and the background bulk (PNG) of the granulocyte data generated for this evaluation.
 }
 \label{fig:coverage-dists}
\end{figure}

@@

@@

SCAN-SNV \citep{luquette_identification_2019} was installed via the install instruction in the repository on 2019-10-30\footnote{
 This are the install instructions at the repository state at \url{https://github.com/parklab/scan-snv/blob/62b7dcbc21e6cac64dc384abdfa708120e33819c/README.md}.
 This includes the following packages and versions from the custom conda channel \url{https://anaconda.org/jluquette}:
 {\ttfamily r-scansnv~0.1}, {\ttfamily scansnv~0.9}, {\ttfamily r-fastghquad~1.0} and {\ttfamily natefoo-slurm-drmaa~1.2.0}.
 Further, it requires manual registration of the java executable for {\ttfamily gatk~3.8}.
} and patched locally to make it run.
The pipeline was then run with parameters set as recommended in the software repository {\ttfamily README.md} and the command line help, with the only changes made for a better comparability with the other tools.
Namely, we tried to call as sensitively as possible by setting: {\ttfamily  --min-sc-alt~1 --min-sc-dp~1 --min-bulk-dp~1}.
However, the default pipeline only provides calls for somatic variants (variants only found in single cells, not in the bulk, as opposed to all the present alternative alleles), heavily filtering candidate sites before calling.
We thus attempted a fairer alternative allele calling comparison by modifying the pipeline to remove the filter that eliminates alternative alleles also seen in the bulk sample and the filter that eliminates known dbsnp variation\footnote{
 I.e.~we edited the snakemake {\ttfamily rule scansnv\_somatic\_sites} to exclude the following requirements:
 {\ttfamily tab[,bulk.alt] == 0 \& tab[,bulk.idx] == '0/0' \& tab\$dbsnp == '.'}.
}.
The different filter thresholds used represent the {\ttfamily --fdr} parameter, which---according to the authors---does not formally control the false discovery rate \citep{luquette_identification_2019}.\\

@@

For SCIPhI (version {\ttfamily 0.1.4}, as available on bioconda), we ran two different modes:
\begin{enumerate*}
 \item The default parameters of the command line tool ({\ttfamily default}).
 \item Fewer iterations ({\ttfamily -l~400000}, because of the excessive run time) and more {\ttfamily sensitive} by turning off all heuristic filters ({\ttfamily --cwm~1 --mnp~1 --ms~1 --bns~0 --bnc~0 --ncf~0 --mnc~1}).
\end{enumerate*}
On the whole exome dataset with five cells, SCIPhI ran for three weeks for each of these modes.
On the whole genome dataset with two cells, an initial run of SCIPhI was killed due to a server problem after running for more than eight weeks.
An attempt to restart from intermediate data using the {\ttfamily --il} command line parameter had the model estimations escalating towards unrealistic values, and the crash was thus left unrecoverable.
A rerun on a server with more performant CPUs led to the sensitive mode finishing after 5 weeks and the default mode after 7.5 weeks.
As no option for filtering or thresholding based on sensitivity vs. specificity is available, only two unfiltered data points are given in the respective plots (Figure~\ref{fig:alt-calling_prec-rec_global}).

@@

\begin{figure}[!tpb]
 \begin{minipage}{.36\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.55\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher. 
In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
SCAN-SNV achieves up to .992 precision, but only .0001 recall.
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
This dataset is considerably more challenging than the first one.
See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .176, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.

>>>>

\documentclass[authoryear,preprint,11pt]{scrartcl}
\addtokomafont{disposition}{\rmfamily}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{url}
\usepackage{epsfig}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

@@

\begin{figure}[tbp]
 \includegraphics[width=.7\linewidth]{figs/Event_space.pdf}\\
 \caption{
  It depicts single cell events at a particular genomic site, defined by combinations of ProSolo's maximum likelihood estimates for possible true alternative nucleotide frequencies in single cell sequencing ($\tilde{\theta}_s$) and bulk sequencing samples ($\tilde{\theta}_b$) in our probabilistic model.
  The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\tilde{\theta}_b$ (bottom panel).
  A tabular representation of the same event definitions is given in Table~\ref{tab:Events} below (this is a larger copy of Figure~\ref{fig:prosolo_alt-calling}D in the main manuscript, reproduced  here for easier cross-reference with that table).\footnotesize\newline
  ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference
  }
 \label{fig:event-space}
\end{figure}

\begin{center}
 \begin{table}[tbp]
  \caption{
   Single cell events at a particular genomic site are defined by combinations of ProSolo's likelihood density estimates of the possible true alternative nucleotide frequency ranges in single cell sequencing ($\tilde{\theta}_s$) and bulk sequencing samples ($\tilde{\theta}_b$) in our probabilistic model.
   A graphical representation of the same event definitions is given in Figure~\ref{fig:event-space}.
  }
  \label{tab:Events}
  \renewcommand{\arraystretch}{0.9}

  \begin{tabular}{p{22ex}p{12ex}p{12ex}p{33ex}}
   \toprule
   single cell event & $\tilde{\theta}_s(x)$   & $\tilde{\theta}_b(x)$ & implicit single cell genotype \\
   \toprule
   $\boldsymbol{E}_{\text{hom ref}}$      & $0$             & $[0,\frac12)$ & homozygous ref                \\
       \midrule
   $\boldsymbol{E}_{\text{ADO to ref}}$   & $0$             & $[\frac12,1)$ & heterozygous                  \\
         \midrule
   $\boldsymbol{E}_{\text{err ref}}$      & $\{0,\frac12\}$ & $1$           & homozygous alt                \\
         \midrule
   $\boldsymbol{E}_{\text{err alt}}$      & $\{\frac12,1\}$ & $0$           & homozygous ref                \\
         \midrule
   $\boldsymbol{E}_{\text{het}}$          & $\frac12$       & $(0,1)$       & heterozygous                  \\
         \midrule
   $\boldsymbol{E}_{\text{ADO to alt}}$   & $1$             & $(0,\frac12]$ & heterozygous                  \\
       \midrule
   $\boldsymbol{E}_{\text{hom alt}}$      & $1$             & $(\frac12,1]$ & homozygous alt                \\
   \bottomrule
   \multicolumn{4}{l}{\footnotesize ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous,}\\
   \multicolumn{4}{l}{\footnotesize hom -- homozygous, ref -- reference} \\
   \bottomrule
  \end{tabular}
 \end{table}
\end{center}

Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell and the bulk sequencing sample, given the respective data.
All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequency likelihood density estimates $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
We then define mutually exclusive sub-spaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$, providing a set of events $\mathfrak{E}$.
For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:event-space}, lower panel).
At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
For the above example ($\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$), this means that we can resolve a contradiction between the likelihood estimate at $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and the likelihood density estimate across $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
Based on our assumption, we decide to trust the bulk sample over the single cell sample, meaning that the bulk is a mixture of a maximum of two sub-populations, and can only contain heterozygous and homozygous alternative cells, rendering a homozygous reference single cell impossible and thus the result of an allele dropout.

Further assuming that there is no informative prior information about different allele frequencies (i.e. assuming a flat prior across all possible allele frequencies in both samples), we can then calculate the likelihood of such an event as the product of the likelihoods of the allele frequency ranges in the two samples.
For $\boldsymbol{E}_{\text{ADO to ref}}$---and using Equation \ref{eq:thetas-likelihood} to calculate the likelihoods of the allele frequency ranges---this for example becomes:

@@

\begin{figure}[!tpb]
 \begin{minipage}{\linewidth}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_coverage_dist.pdf}
 \end{minipage} \\
 \begin{minipage}{\linewidth}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_coverage_dist.pdf}
 \end{minipage}
 \caption{
 Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples (continued on following page).
 Panel \textbf{A} shows the coverage of the single cells (IL-11, IL-12), the kindred ground truth clone (IL-1c) and the more distant background bulk (Clones C1, C2, and C3 joined into one sample) of the dataset from \cite{dong_accurate_2017}.
 Panel \textbf{B} shows the coverage of the single cells (PAG1, PAG2, PAG5, PAG9, PAG10) and the background bulk (PNG) of the granulocyte data generated for this evaluation.
 }
 \label{fig:coverage-dists}
\end{figure}

\begin{figure}[!tpb]
 \ContinuedFloat
 \begin{minipage}{\linewidth}
  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_coverage_dist.pdf}
 \end{minipage}
 \caption{
 Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples (continued from previous page).
 Panel \textbf{C} shows the coverage of the single cells (a1 to a8, h1 to h8, and TNBC-n1 to TNBC-n16), and the bulk samples (TNBC-Pop-Tumor and TNBC-Pop-Normal) that were used as ground truth for the normal single cells (with validated somatic variants added) and as ground truth for the tumor single cells (with validated somatic variants removed), respectively \cite{wang_clonal_2014}.
 }
 \label{fig:coverage-dists}
\end{figure}

@@

\textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
For each, 16 cells and a bulk sample were whole exome sequenced.
For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
Depending on the single cell, we used two different ground truths for two separate evaluations:
\begin{enumerate}
    \item For the tumor single cells, we constructed a ground truth from normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
    Thus, similar caveats as for the germline ground truth in the granulocyte dataset apply.
    To mitigate this to a certain degree, we added the 373 clonal validated somatic variants, as these are expected to be present in all tumor single cells.
    \item For the normal single cells, we constructed a ground truth from the tumor bulk sample (as the normal bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
    To reduce the presence of false-positive somatic variants in this ground truth, we removed all 467 validated somatic variants (clonal and subclonal) from the ground truth.
\end{enumerate}
Further, we determined the of the validated somatic variants for all software run on this dataset.

@@

SCAN-SNV \citep{luquette_identification_2019} was installed via the install instruction in the repository on 2019-10-30\footnote{
 This are the install instructions at the repository state at \url{https://github.com/parklab/scan-snv/blob/62b7dcbc21e6cac64dc384abdfa708120e33819c/README.md}.
 This includes the following packages and versions from the custom conda channel \url{https://anaconda.org/jluquette}:
 {\ttfamily r-scansnv~0.1}, {\ttfamily scansnv~0.9}, {\ttfamily r-fastghquad~1.0} and {\ttfamily natefoo-slurm-drmaa~1.2.0}.
 Further, it requires manual registration of the java executable for {\ttfamily gatk~3.8}.
} and patched locally to make it run.
The pipeline was then run with parameters set as recommended in the software repository {\ttfamily README.md} and the command line help, with the only changes made for a better comparability with the other tools.
Namely, we tried to call as sensitively as possible by setting: {\ttfamily  --min-sc-alt~1 --min-sc-dp~1 --min-bulk-dp~1}.
However, the default pipeline only provides calls for somatic variants (variants only found in single cells, not in the bulk, as opposed to all the present alternative alleles), heavily filtering candidate sites before calling.
We thus attempted a fairer alternative allele calling comparison by modifying the pipeline to remove the filter that eliminates alternative alleles also seen in the bulk sample and the filter that eliminates known dbsnp variation\footnote{
 I.e.~we edited the snakemake {\ttfamily rule scansnv\_somatic\_sites} to exclude the following requirements:
 {\ttfamily tab[,bulk.alt] == 0 \& tab[,bulk.idx] == '0/0' \& tab\$dbsnp == '.'}.
}.
The different filter thresholds used represent the {\ttfamily --fdr} parameter, which---according to the authors---does not formally control the false discovery rate \citep{luquette_identification_2019}.
For the third dataset, added following a suggestion from peer-review, we had to recreate the conda environment for SCAN-SNV, following newer advice in the issues of the repository\footnote{see: \url{https://github.com/parklab/scan-snv/issues/4##issuecomment-596296645}}.
While this allowed us to run the pipeline, the custom rule for ground truth comparison ran into unsolvable version conflicts.
After multiple days of debugging attempts, we decided to exclude SCAN-SNV from the analysis of this dataset, also given its performance on the other datasets.\\

@@

For SCIPhI (version {\ttfamily 0.1.4}, as available on bioconda), we ran two different modes:
\begin{enumerate*}
 \item The default parameters of the command line tool ({\ttfamily default}).
 \item Fewer iterations ({\ttfamily -l~400000}, because of the excessive run time) and more {\ttfamily sensitive} by turning off all heuristic filters ({\ttfamily --cwm~1 --mnp~1 --ms~1 --bns~0 --bnc~0 --ncf~0 --mnc~1}).
\end{enumerate*}
On the whole exome dataset with five cells, SCIPhI ran for three weeks for each of these modes.
On the whole genome dataset with two cells, an initial run of SCIPhI was killed due to a server problem after running for more than eight weeks.
An attempt to restart from intermediate data using the {\ttfamily --il} command line parameter had the model estimations escalating towards unrealistic values, and the crash was thus left unrecoverable.
A rerun on a server with more performant CPUs led to the sensitive mode finishing after 5 weeks and the default mode after 7.5 weeks.
On the third dataset, added after a suggestion during peer-review, we only ran SCIPhI with default parameters and reduced the iterations to {\ttfamily -l 80000} (and the number of iterations for the mutation to cell assignment to {\ttfamily --ls 40000}) to finish analysis in a reasonable time-frame.
No guidance is given for those parameters, but the optimization score seemed to have mostly settled with this amount iterations.
With these parameters, this analysis finished within a week for both the tumor cell and the normal cell datasets of \cite{wang_clonal_2014}.
As no option for filtering or thresholding based on sensitivity vs. specificity is available from SCIPhI itself, we parsed output with custom scripts to extract what must be a posterior probability of the presence of a variant at a particular site into a format that ProSolo's false discovery rate control mechanism can handle.
This is the value in the last (fourth) FORMAT field, which according to the header annotation is either the GQ or the PL field (five FORMAT fields are specified, but only four values per sample are provided).
Judging from the source code\footnote{For the full relevant source code, see here: \url{https://github.com/cbg-ethz/SCIPhI/blob/34975f7050f86325f361c58e8146883af4a2d289/src/output.h##L202-L216}}, this field contains a PHRED scaled posterior probability of the site not being a variant\footnote{This is 1 minus a value that i is used for determining genotypes by SCIPhI with a fixed cutoff of 0.95 a few lines above: \url{https://github.com/cbg-ethz/SCIPhI/blob/34975f7050f86325f361c58e8146883af4a2d289/src/output.h##L205}}.
We parsed this into the probability of a variant being present and passed it into {\ttfamily prosolo control-fdr} to attempt controlling the false discovery rate (Figures~\ref{fig:alt-calling_prec-rec_global} and \ref{fig:FDR-ground-truth-vs-theoretical}).

For scVILP (has no versioning\footnote{We used the repository at this commit: \url{https://github.com/mae6/scVILP/tree/ef38bc422d44daafe043c7dcb28112a1f1ee2fe6}}; install instructions had to be amended\footnote{Pull request to amend install instructions: https://github.com/mae6/scVILP/pull/4} and manually fixing parsing and dependency issues\footnote{Pull request to allow handling of mitochondrial chromosome: \url{https://github.com/mae6/scVILP/pull/1}; Pull request to fix impossible import of {\ttfamily cdecimal} by upgrading to {\ttfamily decimal} \url{https://github.com/mae6/scVILP/pull/2}}), we tried two modes:
\begin{enumerate}
    \item Default parameters as suggested in the README of the software: \url{https://github.com/mae6/scVILP#identify-the-candidate-loci}.
    \item A more sensitive identification of candidates by turning off heuristic filters ({\ttfamily -ms 1 -nmc 1}) during the execution of {\ttfamily loci\_identify.py}.
\end{enumerate}
However, scVILP was not able to handle any of our three benchmarking datasets, with the Gurobi solver that it uses always crashing with an out of memory error, even when provided with more than a Tebibyte of RAM.
We documented details of runs and crashes for the developers of the tool in an issue in their code repository: \url{https://github.com/mae6/scVILP/issues/5}.

@@

\begin{figure}[!tpb]
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.48\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.9\linewidth}
  \includegraphics[height=42ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline
  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily prosolo --fdr}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher. 
In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
SCAN-SNV achieves up to .992 precision, but only .0001 recall.
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants\footnote{E.g. see this comment in the issues of the respective repository: \url{https://github.com/parklab/scan-snv/issues/5##issuecomment-628107370}}.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
This dataset is considerably more challenging than the first one.
See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf}
 \caption{
 Cell-specific $F_1$ scores for different filtering threshold levels for the whole genome dataset.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:F_1-scores_Dong2017}
\end{figure}
